### 2011-02-09 Documentation/atropos/atropos.tex
ENH: minor changes to response and atropos wrt wu's reading.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@772 fd2da0a4-d968-48df-9cb5-d8c496287cbf

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i |
    x_{\mathcal{N}_i}\right) 
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right) 
\end{align}
with $Z$ a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right)
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            0 & \text{if } x_i = x_j \\
                            1 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of
the MRF prior on the segmentation solution.  
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more
heavily than distant sites.

@@

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alternative to Boykov's solution.
Furthermore, one may automate the initialization process.  
{\color{blue}{We provide this capability to allow the user to implement
    an interactive editing and segmentation loop.  The user may
    run Atropos with sparse manual label guidance, evaluate the
    results, update the manual labels and repeat until achieving the desired
    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}}

\begin{comment}
{
\subsubsection{Distance Prior for Label Propagation}

@@

For each iteration, a random ordering of the
set $\{1,\ldots,C\}$ is created which prescribes the order in which the

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  
% The prior label image initialization is competitive with but inferior to K-means.  This is
%  sensible in that the prior labels image outcome is sensitive to the
%  user choices made.  In this study, we labeled the brain web template
%  very sparsely, quickly and crudely and did not expect highly
%  accurate results.  
  The legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4-assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20-subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=0.5$, the priors are maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in Figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

@@

We first convert each of the 69 labels within the original evaluation
dataset to an individual image.  The remaining steps,
summarized briefly, are the same for each of the 19 subjects.  We
select one subject as an unlabeled target.  The other 18 datasets are
then mapped (as in the script above) to that subject.  We then deform,
individually, the 69 $\times$ 18 label images to the unlabeled
subject.  The label probability map is gained by averaging the 18
deformed images associated with each label.  We repeat this for each
subject.  The following parameters are the most relevant to this discussion:
{\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
-m [0.1,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
Dice overlap, are shown in Figure~\ref{fig:hammers}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small The figure compares the Dice
overlap results from Atropos versus the raw results from majority
voting for each of 68 neuroanatomical regions and, in addition, the
unlabeled portions of the brain from the Hammers evaluation dataset.
We evaluated Atropos via N-fold cross-validation and employed
PriorProbabilityImages for each class where probabilities are gained
by averaging mapped subject labels. The color coding highlights those
regions that have the highest (yellow) and lowest (pink) improvement.
The significance of the improvement, measured by pairwise T-test, is
also shown as is a trinary coding of that improvement as: $+$
significant improvement, $-$ performance reduction, $\sim$ no change.}
\label{fig:hammers}
\end{center}
\end{figure}

>>>>

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}\label{eq:cond}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i |
    x_{\mathcal{N}_i}\right) 
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right) 
\end{align}
with $Z$ a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in
    \mathcal{N}_i} V_{ij}( x_i, x_j ) \right) \nonumber
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            0 & \text{if } x_i = x_j \\
                            1 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of
the MRF prior on the segmentation solution.  
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more
heavily than distant sites.

@@

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alternative to Boykov's solution.
Furthermore, one may automate the initialization process.  
{\color{blue}{We provide this capability to allow the user to implement
    an interactive editing and segmentation loop.  The user may
    run Atropos with sparse manual label guidance, evaluate the
    results, update the manual labels and repeat until achieving the desired
    outcome.  This processing loop may be performed easily with, e.g., ITK-SNAP.}}
\begin{comment}
{
\subsubsection{Distance Prior for Label Propagation}

@@

For each iteration, one has the option to permute the
set $\{1,\ldots,C\}$ which prescribes the order in which the

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  
% The prior label image initialization is competitive with but inferior to K-means.  This is
%  sensible in that the prior labels image outcome is sensitive to the
%  user choices made.  In this study, we labeled the brain web template
%  very sparsely, quickly and crudely and did not expect highly
%  accurate results.  
  The legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4-assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20-subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=0.5$, the priors are maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in Figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.pdf}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.pdf}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

@@

We first convert each of the 69 labels within the original evaluation
dataset to an individual image.  The remaining steps,
summarized briefly, are the same for each of the 19 subjects.  We
select one subject as an unlabeled target.  The other 18 datasets are
then mapped (as in the script above) to that subject.  We then deform,
individually, the 69 $\times$ 18 label images to the unlabeled
subject.  The label probability map is gained by averaging the 18
deformed images associated with each label.  We repeat this for each
subject.  The following parameters are the most relevant to this discussion:
{\ttfamily  -i  PriorProbabilityImages[69,label\_prob\%02d.nii.gz,0.5] 
-m [0.2,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
Dice overlap, are shown in Figure~\ref{fig:hammers}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small The figure compares the Dice
overlap results from Atropos versus the raw results from majority
voting for each of 68 neuroanatomical regions and, in addition, the
unlabeled portions of the brain from the Hammers evaluation dataset.
We evaluated Atropos via N-fold cross-validation and employed
PriorProbabilityImages for each class where probabilities are gained
by averaging mapped subject labels. The color coding highlights those
regions that have the highest (yellow) and lowest (pink) improvement.
The significance of the improvement, measured by pairwise T-test, is
also shown as is a trinary coding of that improvement as: $+$
significant improvement, $-$ performance reduction, $\sim$ no change.}
\label{fig:hammers}
\end{center}
\end{figure}
